---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

The research dataset zip file should be in the current working directory or it will be downloaded and unzipped to the current working directory from the [webpage][1].


```{r load echo=TRUE}

setwd("~/R Projects/Coursera/5 - Reproducible Research/RepData_PeerAssessment1")
rm(list=ls())

# Load packages
library(dplyr)
library(ggplot2)

# Download uri.
data.url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
data.url <- gsub("%2F", "/", data.url)

file.zip <- "activity.zip"
file.data <- "activity.csv"

# Download file if missing.
if (!file.exists(file.zip)){
  message("Downloading file into: ", getwd())
  download.file(data.url, file.zip, method = "auto", mode="wb") # Don't use curl on Windows.
  unzip(file.zip)
}

# Load into df and set correct column types. Remove NA's.
activity <- read.csv(file.data) %>%
  mutate(date = as.Date(date)) %>%
  mutate(steps = as.integer(steps)) %>%
  mutate(interval = as.integer(interval))


```

## What is mean total number of steps taken per day?

Before creating a histogramm remove NA values and group dataframe by date. Apply
the sum() function to $steps.

```{r part1_calc echo=TRUE}
summ.byDate <- activity %>%
  na.omit() %>% # 15264 left
  group_by(date) %>%
  summarise(
    steps = sum(steps))

```

Draw a histogram for steps.

```{r part1_hist echo=TRUE}
hist(summ.byDate$steps,
     col = "aquamarine3",
     main="Steps per day",
     xlab="Steps")

```

Calculate the mean and median for values.

```{r part1_MeanMed echo=TRUE}
steps.mean <- mean(summ.byDate$steps)
steps.median <- median(summ.byDate$steps)
```

The *mean* total number of steps taken per day is `r steps.mean` and the 
*median* is `r steps.median`.

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?



[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip
